---
title: "STAT545 Assignment 4"
author: "Harper Cheng"
date: "10/2/2019"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries and dataset
```{r, warning=FALSE, message=FALSE}
library(gapminder)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
```


# Exercise 1: Univariate Data Reshaping

## Univariate Option 1

### 1.1 Reshaping
Make a tibble with one row per year, and columns for life expectancy for two or more countries.
```{r}
dat1 <- gapminder %>% 
  pivot_wider(id_cols=year, 
              names_from=country,
              values_from=lifeExp)
dat1 %>% DT::datatable()
```

### 1.2 ggplot
Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.
```{r}
dat1 %>%
  ggplot(aes(x=Afghanistan, y=Albania)) +
  geom_point(size=3, color='blue') +
  labs(title='A Scatterplot of Life Expectancy for Albania and Afghanistan') +
  theme_bw()
```

### 1.3 Re-lengthening 
Re-lengthen the data.
```{r}
dat1 %>% 
  pivot_longer(cols=-year,
               names_to="country",
               values_to="lifeExp") %>% 
  arrange(country) %>% 
  DT::datatable()
```


# Exercise 2: Multivariate Data Reshaping

## Multivariate Option 1

### 2.1 Reshaping 
Make a tibble with one row per year, and columns for life expectancy and GDP per capita for two or more countries.
```{r}
dat2 <- gapminder %>% 
  filter(country %in% c("Afghanistan", "Albania")) %>% 
  pivot_wider(id_cols=year,
              names_from=country,
              names_sep="_",
              values_from=c(lifeExp, gdpPercap))
dat2 %>% DT::datatable()
```

### 2.2 Re-lengthening 
Re-lengthen the data.
```{r}
dat2 %>% 
  pivot_longer(cols=-year,
               names_to=c(".value", "country"),
               names_sep="_") %>% 
  DT::datatable()
```


# Exercise 3: Table Joins

*Read in the made-up wedding guestlist and email addresses using the following lines. Then, complete the following tasks using the tidyverse (tidyr, dplyr, …). No need to do any pivoting – feel free to leave guest in its current format.*
```{r, include=FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

1. For each guest in the guestlist (`guest` tibble), add a column for email address, which can be found in the `email` tibble.
```{r}
email_list <- email%>% 
  separate_rows(guest, sep=", ") %>% 
  rename(name=guest)

guest %>% 
  left_join(email_list, by="name")
```

2. Who do we have emails for, yet are not on the guestlist?
```{r}
email_list %>% 
  anti_join(guest, by="name")
```

3. Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).
```{r}
email_list %>% 
  full_join(guest, by="name")
```



