---
title: "STAT545_Assignment 1"
author: Xiaolan Harper Cheng
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this R Markdown document, the dataset 'Gapminder' is briefly analyzed. 

## Data Analysis

Let us first load the dataset and explore the structure of it. It is described in
the R documentation that 'gapminder' is an excerpt of the Gapminder data on life
expectancy, GDP per capita, and population by country. 

```{r load data}
library(gapminder)
?gapminder
str(gapminder)
```

It is suggested by the internal structure that 'gapminder' is a data frame with 
1704 observations and 6 variables which are country, continent, year, life expectancy, population, and GDP per capita. 

Japanese people have the longest life expectancy, while Kuwait is the richest country.

```{r }
gapminder$country[gapminder$lifeExp==max(gapminder$lifeExp)]
gapminder$country[gapminder$gdpPercap==max(gapminder$gdpPercap)]
```

Next, the relationship between life expectancy and GDP per capita are analyzed. 
The scatter plot indicates that people in wealthier countries generally have longer
life spans.

```{r warning=FALSE}
attach(gapminder)
agg <- aggregate(gapminder, by=list(country), FUN=mean)
scatter.smooth(y=agg$lifeExp, x=agg$gdpPercap, 
               main="Relationship b/w life expectancy and GDP per capita", 
               ylab="Life Expectancy", xlab="GDP per capita")
```


